Optical Music Recognition for LilyPond File Generation

### 2024-05-11 papers/cs444/paper-cs444.tex
paper double-checked

In the world of music or audio transcription/arrangement, the complexity of sound and score data has allowed human performance to remain the current state-of-the-art. 
Several factors contribute to this circumstance: the number of potential audio and score file types, ambiguity on how performance qualities are notated, and overall inconsistencies between recordings and their respective scores. 
Machine learning models, in turn, have been a crucial step towards reducing these inconsistencies. 
Their ability to learn the nonlinearities and artistic qualities that otherwise plague audio computation have allowed for noteworthy advancements such as the WaveNet~\cite{oord2016wavenet}. 
The remaining issue in the process of sound generation is the amount of data available to train with. 
Worthwhile audio data remains difficult to collect due to its large size and potential copyright issues. 
In particular, trying to condition off another medium is incredibly difficult as current datasets lack the correlations necessary to streamline the audio transcription process. 
Datasets such as MAESTRO ~\cite{hawthorne2018enabling} are close to ideal results, but some intermediate steps are needed to learn correlations between audio files and musical scores.

@@

Given the vast complexity of western music notation and intricacies in notation/time alignment, the proposed model has a pessimistic approach. 
That is, given a small subset of musical data (\textit{among all notated music}) and some model, results can be expected to only match expected outputs through one or more of: pitch, rhythm, duration, and overall formatting.
Existing research in OMR supports these expectations. First, several digital notation systems exist for music, including:

\begin{itemize}
	\item Standards like \textit{MusicXML}, \textit{Lilypond} and \textit{MIDI},
	\item Software-specific formats from \textit{Musescore}, \textit{Sibelius} and \textit{Finale},
	\item \textit{KERN} from the Humdrum tool-set ~\cite{contreras2023omrpian},
	\item Mayer, et al.'s \textit{Linearized MusicXML} ~\cite{mayer2024practical}, and
	\item Contreras, et al.'s untitled "end-to-end OMR" encoding language ~\cite{contreras2023omrcnn}.
\end{itemize}

@@

Despite low research expectations, I believe that a few questions can be posed and answered for the purpose of bounding requirements on a larger, all-encompassing Image-to-MIDI model:

@@

\section{Background}
\subsection{Optical Music Recognition}
Current research in Optical Music Recognition (OMR) can be put in one of two categories:

@@

On the other hand, Music Transcription entirely of the Symbol Recognition problem, (iteratively for some amount of sheet music) as well as problems with correlating recognized symbols in sequence.
On its own, Symbol Recognition could recognize every component of a piece of music, but these results lack the ordered context necessary for transcription.
State-of-the-art transcription through machine learning, as a result, has been dominated by the Convolutional Recurrent Neural Network (CRNN). 
Unlike CNNs, CRNNs make use of recurrent layers, (such as bidirectional LSTM layers in Shi, et al. ~\cite{shi2015endtoend}) in order to contextualize sequences of data forwards and backwards.
These layers, along with attention cells focused on deriving pitch, rhythm and durational context, allow for CRNNs to recognize and output ordered sequences of logical musical data.

@@

At the music level, rhythmic musical symbols are notated according to their pitches and relative time duration.
pitches are all characters ${a, b, c, d, e, f, g, r}$, where $r$ is a "rest" meaning no pitch occurs.
To ascend or descend in pitch beyond a single musical "octave," the $'$ and $,$ characters are appended to indicate one or more octave ascendings or descendings, respectively.
Pitches are also proceeded by a rhythmic value representing its duration in time.
These values are typically powers of two, (but can technically be any positive floating-point value), and they dictate how long a pitch is played with respect to the piece's time signature.
For example, a score with time signature = 2/4 indicates that each measure has two beats, and each beat is the length of a quarter note (hard-coded in the denominator).
In this case, the line $"c4 \quad d4"$ would represent two quarter (4) notes or a single measure in the provided time signature.
Additionally, notes without defined durational values take on the previous note's duration in a line, so a series of equal-duration notes is represented by one durational value.
Finally, separations between measures are made with $"| \% n"$, marking the end of measure $n$.
Figure~\ref{fig:lilypond-example} compares the rendering of a single staff of music with its relative Lilypond notation.
\footnote{The time signature $C$ or "common-time" is another common way of writing 4/4.}

\begin{figure}
	\begin{subfigure}
		\centering
		\includegraphics[width = .8\linewidth]{./figures/lilypond-snippet.png}
	\end{subfigure}
	~
	\begin{subfigure}
		\centering
		\includegraphics[width = .8\linewidth]{./figures/lilypond-code.png}
	\end{subfigure}
	\caption{Lilypond (.ly) code and rendering}
	\label{fig:lilypond-example}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width = .8\linewidth]{./figures/crnn_arch.png}
	\caption{General CRNN architecture, as modeled by Shi, et al. ~\cite{shi2015endtoend}}
	\label{fig:crnn-arch}
\end{figure}

@@

\subsection{Quantization Rendering}
Similar to simple transformations performed on images, (rotations, coloring, scaling), output for sheet music determined by technical, but musical variables that don't affect the "performance" of the composition.
In particular, when score images are rendered from the Bach lilypond files, (trivially translated from the Bach MIDI Index~\cite{bach_midi}), there is a required "quantization" value which affects the rhythms and durations visually.
This quantization, designated $Q \in \mathbb{Z}$, represents what specific note duration counts as a "beat" within a given score. 
Figure ~\ref{fig:table-duration} lists the most common duration mappings that occur within this experiment.

\begin{figure}
	\centering
	\includegraphics[width = .8\linewidth]{./figures/notevalues.jpeg}
	\caption{Table of common duration-integer mappings, from Francis Hamzagic ~\cite{hamzagic_2018}}
	\label{fig:table-duration}
\end{figure}

@@

\begin{figure}
	\centering
	\begin{subfigure}
		\centering
		\includegraphics[width = .8\linewidth]{./figures/c1.png}
	\end{subfigure}
	~
	\begin{subfigure}
		\centering
		\includegraphics[width = .8\linewidth]{./figures/c32.png}
	\end{subfigure}
	\caption{Levels of Quantization $Q = [1, 32]$ in Rendering}
	\label{fig:quantization}
\end{figure}

@@

\subsection{Model Criticism (RQ3)}
My CRNN implementation, along with the CRNN model for which my research references, are not without active barriers preventing consistent musical generalization.
In staying consistent with my description of the Lilypond language, my critiques against CRNNs are categorized by what further work should be done in order for music-level and document-level generalizations to occur.

@@

\subsubsection{Document-level Generalization}
At the document level, the greatest issue plaguing OMR research is the lack of a formatting language standard, or lack of similarity between all formatting languages.
Every formatting language has separate means of conveying equivalent musical information based on the user's needs, (i.e. MIDI for real-time data, MusicXML for web-related music, Lilypond for polished layouts).
However, as the needs of users vary in the numerous niches of music production, technology and composition, digital music encoding and datasets slowly slow the possibility of standardization.
While MIDI and MusicXML are currently the most popular for datasets, both lack consistency in rendering sheet music across all digital notation platforms.
I believe consistency on this front is crucial for reducing variability in musical machine learning datasets. In particular, sheet music should "look the same" regardless of the formatting language it is imported from.

>>>>

In the world of music or audio transcription/arrangement, the complexity of sound and score data has allowed human performance to remain the current state-of-the-art. 
Several factors contribute to this circumstance: the number of potential audio and score file types, ambiguity on how performance qualities are notated, and overall inconsistencies between recordings and their respective scores. 
Machine learning models, in turn, have been a crucial step towards reducing these inconsistencies. 
Their ability to learn the nonlinearities and artistic qualities that otherwise plague audio computation have allowed for noteworthy advancements such as WaveNet~\cite{oord2016wavenet}. 
The remaining issue in the process of sound generation is the amount of data available to train with. 
Worthwhile audio data remains difficult to collect due to its large size and potential copyright issues. 
In particular, trying to condition off another medium is incredibly difficult as current datasets lack the correlations necessary to streamline the audio transcription process. 
Datasets such as MAESTRO ~\cite{hawthorne2018enabling} are close to ideal results, but some intermediate steps are needed to learn correlations between audio files and musical scores.

@@

Given the vast complexity of classical western music notation and intricacies in notation/time alignment, the proposed model maintins a rather pessimistic approach. 
That is, given a small subset of musical data (\textit{among all notated music}) and some model, results can be expected to only match expected outputs through one or more of: pitch, rhythm, duration, and overall formatting.
Existing research in OMR supports these expectations. First, several digital notation systems exist for music, including:

\begin{itemize}
	\item Standards like \textit{MusicXML}, \textit{Lilypond} and \textit{MIDI},
	\item Software-specific formats from \textit{Musescore}, \textit{Sibelius} and \textit{Finale},
	\item \textit{KERN} from the Humdrum tool-set ~\cite{contreras2023omrpiano},
	\item Mayer, et al.'s \textit{Linearized MusicXML} ~\cite{mayer2024practical}, and
	\item Contreras, et al.'s untitled "end-to-end OMR" encoding language ~\cite{contreras2023omrcnn}.
\end{itemize}

@@

Despite low expectations for training and results, I believe that a few questions can be posed and answered for the purpose of bounding requirements on a larger, all-encompassing Image-to-MIDI model:

@@

\section{Background}
\subsection{Optical Music Recognition}
Current research in Optical Music Recognition (OMR) can be placed in one of two categories:

@@

On the other hand, Music Transcription encompasses the entire Symbol Recognition problem, (iteratively for some amount of sheet music) as well as problems with correlating recognized symbols in sequence.
On its own, Symbol Recognition could recognize every component of a piece of music, but these results lack the ordered context necessary for transcription.
State-of-the-art transcription through machine learning, as a result, has been dominated by the Convolutional Recurrent Neural Network (CRNN). 
Unlike CNNs, CRNNs make use of recurrent layers, (such as bidirectional LSTM layers in Shi, et al. ~\cite{shi2015endtoend}) in order to contextualize sequences of data forwards and backwards.
These layers, along with attention cells focused on deriving pitch, rhythm and durational context, allow for CRNNs to recognize and output ordered sequences of logical musical data.

@@

At the music level, rhythmic musical symbols are notated according to their pitches and relative time duration.
pitches are all characters ${a, b, c, d, e, f, g, r}$, where $r$ is a "rest" meaning no pitch occurs.
To ascend or descend in pitch beyond a single musical "octave," the $'$ and $,$ characters are appended to indicate one or more octave ascendings or descendings, respectively.
Pitches are also proceeded by a rhythmic value representing its duration in time.
These values are typically powers of two, (but can technically be any positive integer), and they dictate how long a pitch is played with respect to the piece's time signature.
For example, a score with time signature = 2/4 indicates that each measure has two beats, and each beat is the length of a quarter note (hard-coded in the denominator).
In this case, the line $"c4 \quad d4"$ would represent two quarter (4) notes or a single measure in the provided time signature.
Additionally, notes without defined durational values take on the previous note's duration in a line, so a series of equal-duration notes is represented by one durational value.
Finally, separations between measures are made with $"| \% n"$, marking the end of measure $n$.
Figure~\ref{fig:lilypond-example} compares the rendering of a single staff of music with its relative Lilypond notation.
\footnote{The time signature $C$ or "common-time" is another common way of writing 4/4.}

\begin{figure}
	\begin{subfigure}
		\centering
		\includegraphics[width = .8\linewidth]{./figures/lilypond-snippet.png}
	\end{subfigure}
	~
	\begin{subfigure}
		\centering
		\includegraphics[width = .8\linewidth]{./figures/lilypond-code.png}
	\end{subfigure}
	\caption{Lilypond (.ly) code and rendering.}
	\label{fig:lilypond-example}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width = .8\linewidth]{./figures/crnn_arch.png}
	\caption{General CRNN architecture, as modeled by Shi, et al. ~\cite{shi2015endtoend}.}
	\label{fig:crnn-arch}
\end{figure}

@@

\subsection{Quantization Rendering}
Similar to simple transformations performed on images, (rotations, coloring, scaling), output for sheet music is determined by both technical and musical constraints that don't affect the "performance" of the composition.
In particular, when score images are rendered from the Bach lilypond files, (trivially translated from the Bach MIDI Index~\cite{bach_midi}), there is a required "quantization" value which affects the rhythms and durations visually.
This quantization, designated $Q \in \mathbb{Z}^+$, represents what specific note duration counts as a "beat" within a given score. 
Figure ~\ref{fig:table-duration} lists the most common duration mappings that occur within this experiment.

\begin{figure}
	\centering
	\includegraphics[width = .8\linewidth]{./figures/notevalues.jpeg}
	\caption{Table of common duration-integer mappings, from Francis Hamzagic ~\cite{hamzagic_2018}.}
	\label{fig:table-duration}
\end{figure}

@@

\begin{figure}
	\centering
	\begin{subfigure}
		\centering
		\includegraphics[width = .8\linewidth]{./figures/c1.png}
	\end{subfigure}
	~
	\begin{subfigure}
		\centering
		\includegraphics[width = .8\linewidth]{./figures/c32.png}
	\end{subfigure}
	\caption{Levels of quantization $Q = (1, 32)$ in rendering, respectively.}
	\label{fig:quantization}
\end{figure}

@@

\subsection{Model Criticism (RQ3)}
My CRNN implementation, along with the CRNN models for which my research references, are not without active barriers preventing consistent musical generalization.
In staying consistent with my description of the Lilypond formatting language, my critiques against CRNNs are categorized by what further work should be done in order for music-level and document-level generalizations to occur.

@@

\subsubsection{Document-level Generalization}
At the document level, the greatest issue plaguing OMR research is the lack of a formatting language standard, or lack of similarity between all formatting languages.
Every formatting language has separate means of conveying equivalent musical information based on the user's needs, (i.e. MIDI for real-time data, MusicXML for web-related music, Lilypond for polished layouts).
However, as the needs of users vary in the numerous niches of music production, technology and composition, digital music encoding and datasets slowly diminish the possibility of standardization.
While MIDI and MusicXML are currently the most popular for datasets, both formats lack consistency in rendering sheet music across all digital notation platforms.
I believe consistency on this front is crucial for reducing variability in musical machine learning datasets. In particular, sheet music should "look the same" regardless of the formatting language it is imported from.
